Parsing Postfix log files

### 2007-04-15 doc/logparser.tex
Added review of other parsers.

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database.  The gathered data can then
    be used to optimise current anti-spam measures, provide a baseline to
    test new anti-spam measures against, or to produce statistics showing
    how effective those measures are.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator, every Real-time Black List returns a
different explanatory message, policy servers may have different messages
depending on the characteristics of the connection, \ldots.  This paper
documents the difficult process of parsing Postfix logs, and presents a
program which parses logs and places the resulting data into a database.
The gathered data can then be used to optimise current anti-spam measures,
provide a baseline to test new anti-spam measures against, or to produce
statistics showing how effective those measures are.  There are numerous
other uses for such data: improving server performance by identifying
troublesome destinations and reconfiguring appropriately or dedicating
transports to those destinations; identifying high volume uses (e.g.
customer newsletters) and restricting those uses to off-peak times; as a
base for billing customers on a shared server; etc.

Section~\ref{background} provides background information about: the
motivation for this project; the Postfix Mail Transport Agent; a short
overview of the parser; the assumptions inherent in the design of the
algorithm and program; and a list of other Postfix log parsers.

@@

XXX COMPARE TO OTHER PARSERS

@@

XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE

@@

XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
THEM FROM WITHIN THE PAPER

>>>>

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible to achieve a high degree of accuracy in understanding
    the logs and reconstructing the actions taken by Postfix to generate
    the logs.  This paper describes the process required, documenting the
    parsing algorithm and rules, explaining the difficulties encountered,
    with reference to an implementation which stores data gathered from the
    logs in an SQL database.  The gathered data can then be used to
    optimise current anti-spam measures, provide a baseline to test new
    anti-spam measures against, or to produce statistics showing how
    effective those measures are.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator; every Real-time Black List returns a
different explanatory message; policy servers may have different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.  customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about: the
motivation for this project; the Postfix Mail Transport Agent; a short
overview of the parser; the assumptions inherent in the design of the
algorithm and program; using a database schema as an API; and a comparison
against other Postfix log parsers.

@@

@@

10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
of the background research for this project.  None of the reviewed parsers
perform the type of advanced parsing and log correlation described here;
all are intended to perform a specific parsing and reporting task, rather
than be a generic parser and leave generating reports from the data to
other programs.  Some save the data extracted to a database but the
majority discard all data once finished running, making historical analysis
impossible.  The other parsers listed all produce a report of greater or
lesser complexity and detail, whereas the program described here doesn't
attempt to produce a report at all; that responsibility is deferred to a
separate program, to be developed later.  The parsing algorithm and program
described here are designed to enable much more detailed log analysis by
providing a stable platform for subsequent programs to develop upon.

XXX DO I NEED TO SAY MORE??

@@

\subsection{Other Postfix log parsers}

\label{other-parsers}

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a "heads up" for potential trouble spots.}
        \url{http://jimsun.linxnet.com/postfix_contrib.html}

    \item [Sawmill Universal Log File Analysis and Reporting] is a
        general purpose commercial product which parses 687 log file
        formats (correct as of 2007/04/15) and produces reports.  I have not
        experimented with it due to its limited data extraction facilities;
        it does have three different sets of data for Postfix (as of
        2007/04/15, one is beta), but they do not appear to be interlinked,
        nor does it save sufficient data for the purposes of this project.
        The source code is available in an obfuscated form only (presumably
        for a fee), and the product is quite expensive, as it requires a
        license for each report which is to be generated. \newline
        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}

    \item [Splunk] XXX INVESTIGATE FURTHER \url{http://www.splunk.com}

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.}  \newline
        \url{http://www.enderunix.org/isoqlog/}

    \item [AWStats] \textit{AWStats is a free powerful and featureful tool
        that generates advanced web, streaming, ftp or mail server
        statistics, graphically.}

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator: the
        data it will extract from an MTA log file is:
        \texttt{time2, email, email\_r, host, host\_r, method, url, code,
        and bytesd.} \newline
        \url{http://awstats.sourceforge.net/} \newline
        \url{http://awstats.sourceforge.net/awstats.mail.html} \newline
        \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}

    \item [Log analyser - throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur.  \newline
        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}

    \item [Anteater] \textit{The Anteater project is a Mail Traffic
        Analyser. Anteater supports currently the logformat produced by
        Sendmail and by Postfix. The tool is written in 100\% C++ and is
        very easy to customize. Input, output, and the analysis are modular
        class objects with a clear interface. There are eight useful
        analyse modules, writing the result in plain ASCII or HTML, to
        stdout or to files.}

        Anteater hasn't been updated since November 2003, and doesn't have
        any English documentation.

        \url{http://anteater.drzoom.ch/}

    \item [Yet Another Advanced Logfile Analyser] uses a plugin based
        system to analyse log files and produce output reports.  The core
        code is merely 91 lines long, as all the parsing and report
        generation is handled by modules.  Using YAALA as a base would be
        only slightly less work as both input and output modules would need
        to be written; it may even be more work to implement the parser
        within the constraints of YAALA.

        \url{http://yaala.org/}

    \item [Logparser/Lire] Lire is another general purpose log parser, but
        is developed under the GNU GPL.  Like other general purpose log
        parsers it doesn't extract enough data from Postfix logs to be
        worthwhile.  Development also seems to have stalled, with only a
        minor update since October 2004.  \newline
        \url{http://logreport.org/lire/}

    \item [Logrep] \textit{Logrep is a secure multi-platform framework for
        the collection, extraction, and presentation of information from
        various log files.}

        The parsing Logrep performs is extremely basic, counting the number
        of different processes executed, from address, to addresses, number
        of recipients, bytes transferred and delay.  \newline
        \url{http://www.itefix.no/phpws/index.php}

\subsection{Postfix resources}

\subsection{SMTP references}

XXX ADD RESOURCES/REFERENCES HERE AND THEN MAKE REFERENCES TO THEM FROM WITHIN THE PAPER
